---
title: "Classes"
output: html_document
---

Object oriented programming is possible in R... somehow...

Classes were introduced into S in version 3 and are called S3 classes. As R is a S dialect the term `S3 Class` has been adopted.

Why is the term `S3` important? Because with S4 an additional class concept has been introduced! Yes, there're two different class concepts
in S/R! To make things worse, the syntax of the two class systems is different. Using S4 classes is different to the way you use common classes found in R packages.

What to use? This depends on who you ask. Some prefer the S4 classes, others, like Google, stick to the S3 classes.

[Google's R Style Guide](https://google-styleguide.googlecode.com/svn/trunk/Rguide.xml)

> avoid S4 objects and methods when possible; never mix S3 and S4 

# S3 Classes

An S3 Class is actually just a common R list with an attribute called `class`.

The simples class
```{r}
object <- list(x=1)
class(object)
class(object) <- "simpleClass"
class(object)
object
```

Of course this is not the best practice to instanciate an object. Using a constructor function should be the way to do it. 

The problem with S3 classes is that it's just a list. You can add and remove elements and you can even `unclass` the list again. It's not type safe, the compiler won't complain when trying to write an unknown member.

The reasons why S3 classes are so popular are:

- it's simple
- it's flexible
- most packages use them

## Constructor function
  
```{r}
powerTool <- function(base) {
    get <- function() base
    set <- function(newBase) base <<- newBase
    sqr <- function() pow(2)
    pow3 <- function() pow(3)
    pow <- function(y) base ^ y
    instance <- list(
        get=get,
        set=set,
        sqr=sqr,
        pow3=pow3)
    class(instance) <- 'powerTool'
    return(instance)
}
```

- methods are simple variable assignments
- at the end of the function we create a list containing all public methods
- the class type is assigned with the `class()` function
- note the new operator `<<-` in the `set()` function
    - this operator is looking in the parent environment for the `myX` variable
    - `<-` would create a local `myX` variable

Usage
```{r}
p <- powerTool(2)
p$sqr()
p$pow3()
```
Note that the `pow()` function is private and not accessible!
```{r, error=TRUE}
p$pow(10)
```

If you just print the variable the ouput is just like the output of a list
```{r}
p
```

We need to implement a `print()` method to change that output. `print()` is a generic function and the implementation is choosen by the class type.

```{r}
print.powerTool <- function(o) cat("PowerTool, base", o$get(), sep = "=")
```

Now we can print the variable again

```{r}
p
```


