---
title: "Functions"
output: html_document
---

In R functions are objects and can be manipulated in much the same way as any other object. Functions (or more precisely, function closures) have three basic components: a formal argument list, a body and an environment

#Function Definition
Functions are written like below (usually in their own file). The value of the last statement gets returned.
```{r}
f <- function() {
  1 + 2
}
class(f)
f()
```
Since Functions are first class objects, a function can return a function.
```{r}
f <- function() {
  g <- function(){
    4 + 4
  }
}
class(f)
a <- f()
class(a)
a()
```
#Arguments
Arguments are written within the brackets
```{r}
pow <- function(x, y){
  x^y
}
pow(2,3)
```
##Default values for Arguments
Arguments can have default values.
```{r}
pow <- function(x, y = 2){
  x^y
}
pow(2,3)
pow(2)
```
##Lazy evaluation of arguments
Arguments are only evaluated when they are needed.
```{r}
f = function(a,b) {
  1 + 2
}
f()
f(1)
f(1,2)
```
##Argument Matching
Arguments can be matched by position or by name (or by a mixture of that)
```{r}
pow <- function(x, y){
  x ^ y
}

pow(2,3)
pow(y = 3, x = 2)
pow(y = 3, 2)
```

That is usefull for functions with long argument lists where one does not want to specify all parameters
```{r}
args(lm)
```

Arguments can also be partially matched
```{r}
pow <- function(myLongParameterNamme, y){
  myLongParameterNamme ^ y
}

pow(y = 3, m = 2)
pow(3, m = 2)
```

The matching order is:

* Exact match by name
* Partial match by name
* Match by position

```{r}
pow <- function(value, value2){
  value^value2
}

pow(value = 2, val = 3)
pow(value2 = 2, 3)
```

##The "..." Argument

... indicates a variable number of arguments. Usually that is used to pass them to another function.
```{r}
pasteWithDashes <- function(...){
  paste(..., sep="-")
}

pasteWithDashes('a', 'b', 'c')
```

Everything after ... argument must be named explicitely.

# Environments
Environments can be thought of as consisting of two things: a frame, which is a set of symbol-value pairs, and an enclosure, a pointer to an enclosing environment. When R looks up the value for a symbol the frame is examined and if a matching symbol is found its value will be returned. If not, the enclosing environment is then accessed and the process repeated. Environments form a tree structure in which the enclosures play the role of parents. The tree of environments is rooted in an empty environment, available through emptyenv(), which has no parent.

So environments are a place to store variables. If ones starts up r and makes an assignment a variable in the global environment is created.
```{r}
rm(list = ls())
a <- 1
ls()
ls(globalenv())
```
##Functions and Environments (Lexical Scoping)
A function's environment is the environment that was active at the time that the function was created. [...] When a function is called, a new environment (called the evaluation environment) is created, whose enclosure is the environment from the function closure. This new environment is initially populated with the unevaluated arguments to the function; as evaluation proceeds, local variables are created within it.

Lets create a function within a function and see what environments are created:

```{r}
rm(list = ls())

f <- function(){
  g <- function(){
  }
}
f()
environment(f)
```

The environment of f is the global environenment.
```{r}
environment(f)
```

The environment of g however is the environment created when f was called.
```{r}
g <- f()
environment(g)
```
The parent of g's environment is f's environment which is the global environment again.
```{r}
parent.env(environment(g))
```

##Lexical Scoping
That now gets interesting together with free variables in functions.
Free variables are variables that arent assigned in the function body and that are not in the attribute list.
```{r}
rm(list = ls())
a <- 2000
b <- 1000
x <- 1
f <- function(){
  a <- 200
  b <- b <- 100
  y <- 2
  g <- function(){
    a <- 20
    x + y + a + b
  }
}
g <- f()
```
The environments do now contain some variables.
```{r}
ls(environment(g))
ls(environment(f))
```
Note that a and b are contained in different environments but with different values
```{r}
environment(g)$a
environment(f)$a
environment(g)$b
environment(f)$b
```
If we now call g, R tries to assign values to the variables for the calculation (x + y + a + b). Therefore it looks through the environments starting with the environment that is created when g is called and if it doesn't find a symbol, it moves up to the parent.
```{r}
g()
```
If it reaches the empty environment, it throws an exception
```{r error=TRUE}
rm(x)
g()
```
#Generic functions and method dispatch
How does the following work?
```{r}
summary(factor(c("big","medium","small","big", "small")))
summary(matrix(1:4, ncol=2, nrow=2))
```
If we look at the function help file, it says "summary is ageneric function...." and if we look at the implementation we see a call to a function UseMethod("summary").
```{r}
summary
```
The help says: Method dispatch takes place based on the class(es) of the first argument to the generic function or of the object supplied as an argument to UseMethod.
So what happens is that R checks the class of the argument and then calls a specific function:
```{r}
summary.factor(factor(c("big","medium","small","big", "small")))
```
We can also write out own generic functions.
```{r}
f.default <- function(x){x + 1}
f.integer <- function(x){x + 2}
f <- function(x){UseMethod("f")}
f(1)
f(1L)
```
