---
title: "Packages"
output: html_document
---

# Package: dplyr

From `?dplyr`

> dplyr provides a flexible grammar of data manipulation. It [...] focussed on tools for working with data frames (hence the d in the name).

dplyr provides a few basic functions:

- select()
- filter()
- arrange()
- manipulate()
- summarize()

## Installation
```R
install.packages("dplyr")
```

```{r, message=FALSE}
library("dplyr")
```

## The problems with standard R dataframes and tables

For R newbies, working with data frames is hard to learn. While simple subsetting is easy much time is spent when trying advanced column or row subsetting and summarizing.

Personally, when solving simple R questions I used to spend a hell lot of time with data frame manipulations. This seems to be general problem. Why I came to that conclusion? Because several people out there wrote data manipulation packages like:

- dplyr
- data.table
- ...

### Example
```{r}
df <- read.csv("data/2014-07-08.csv")
```

Get some information about the data set.
```{r}
head(df)
nrow(df)
```

I'm only interested in some columns, say `package` and `size`... and I'm lost :-)
```{r, error=TRUE}
df[, df$package] 
names(df)
head(df[, 8])
p <- df[,8]
s <- df[,4]
ps <- data.frame(packages=p, size=s)
head(ps)
```

## select()

With `select()` we can select a subset of columns of a data frame.

I'm only interested in some columns, say `package` and `size`... and I'm in heaven!
```{r}
data <- tbl_df(df)
ps <- select(data, package, size)
ps
```

Select a range of columns
```{r}
names(data)
select(data, r_version:ip_id)
```
Same as before but without the coutry
```{r}
select(data, r_version:ip_id, -country)
```
Same as before but in reverse column order
```{r}
select(data, ip_id:r_version, -country)
```

## filter() - Subsetting

Use `filter()` for selecting rows.

```{r}
filter(data, package=="dplyr")
filter(data, package=="dplyr", country=="US")
filter(data, package=="dplyr", country=="US" | country=="IN")
filter(data, !is.na(version))
```


## arrange() - Sorting

Sort by columns by using `arrange()`. For descending order wrap the column name with the `desc()` function.
```{r}
arrange(data, size)
arrange(data, desc(size))
arrange(data, package, desc(version))
```

## mutate() - Add new columns

With `mutate()` it's possible to add new columns.

Calculate the size in MB
```{r}
mutate(data, size_mb = size / 2^20)
mutate(data, size_mb = size / 2^20, size_mb_rounded=round(size_mb, digits = 2))
```

## summarize()

Summarise multiple values to a single value. Summarising by group is supported.

What's the mean size?

```{r}
summarize(data, mean(size))
```

### group_by()

Adds a grouping attribute to the data frame. Summarising on such a data frame will apply the given function on each group.
```{r}
byPackage <- group_by(data, package)
packSum <- summarize(byPackage,
                      count = n(),
                      unique = n_distinct(ip_id),
                      countries = n_distinct(country),
                      avg_bytes = mean(size))
packSum
```

Having this we can solve questions like: Which are the top most 1% packages in the download ranking.
```{r}
q <- quantile(packSum$count, probs = 0.99)
q
topPackages <- filter(packSum, count > q)
topPackages
```
Let's sort the result to get a ranking
```{r}
arrange(topPackages, count)
```

As we can see another data manipulation package is at the top of the download charts!

## Chaining

With the chain operator `%>%` defined in the dplyr package we can reduce the code. The result of a function is the input of the
next function. So we don't have to include the data frame variable in every function call.
```{r}
smallPackages <- data %>% select(package, size, r_version) %>% arrange(package, desc(r_version)) %>%  filter(size < 1000)

smallPackages
```
or a bit more readable
```r
smallPackages <-  
    data %>% 
    select(package, size, r_version) %>% 
    arrange(package, desc(r_version)) %>%  
    filter(size < 1000)
smallPackages
```

# Package: lubridate

Date tools.

## Installation
```{r}
library(lubridate)
```

## Parsing

```{r}
ymd("20141201")
ymd("2014/12/01")
dmy("01122014")
ymd_hms("20141201 10:14:12")
ymd_hms("2014-12-01 10:14:12")
```

##  Get me all the mondays

```{r}
daysOfTheYear <- ymd(20141231) + days(1:365)
head(wday(daysOfTheYear), n=3)
head(wday(daysOfTheYear, label=TRUE), n=3)
daysOfTheYear[wday(daysOfTheYear, label=TRUE) == "Mon"]
```

# Package: base

Some nice functions from the base backage that comes with R.

## grep()/grepl()

As with a Unix CLI we can grep a dataset.

```{r}
grep("2015-02", daysOfTheYear)
grep("2015-(01|02)-01", daysOfTheYear)
grep("2015-(01|02)-01", daysOfTheYear, value = TRUE)
firstOfMonth <- grepl("2015-(01|02)-01", daysOfTheYear)
daysOfTheYear[firstOfMonth]
```