---
title: "Type System"
output: html_document
---


#Data Types
##Vectors
The basic unit of data in R is the vector. (Even a single integer is represented as a vector of length 1)

###Vector Types
R has six basic (‘atomic’) vector types:

####logical
```{r}
vec <- c(TRUE, FALSE)
typeof(vec)
str(vec)
```
####double
```{r}
vec <- c(1, 1.1)
typeof(vec)
str(vec)
```
####integer
```{r}
vec <- c(1L, 2L)
typeof(vec)
str(vec)
```
####complex
```{r}
vec <- c(1i, 1+1i)
typeof(vec)
str(vec)
```
####string (or character)
```{r}
vec <- c('a', 'b')
typeof(vec)
str(vec)
```
####raw
```{r}
vec <- raw(2)
typeof(vec)
str(vec)
```

###Vector Creation
```{r}
c(1,2,3,5,8) #generate a vector with specific elements
1:10 # generate a sequence-vector
vector("numeric", length=10)  # generate a numerical vector of length 10 with default values
```

###Indexing
```{r}
vec <- c('a','b','c','d','e')
```
Vectors can be accessed by index. Beginning with 1.
```{r}
vec[1]
```
The index can also be a vector.
```{r}
vec[3:4]
```
Vectors can also be accessed by boolean vectors.
```{r}
vec[vec == 'c']
```

##Lists
A list is a special type of a vector that can contain different types. (They are Generic Vectors)
```{r}
l <- list(1, "a", TRUE, 1+1i)
l
str(l)
```
###Indexing
One can acess lists similarly to vectors.
```{r}
l <- list(1, "a", TRUE, 1+1i)
l[[1]]
l[[2]]
```
If we assign names to the entries, we can access them by name.
```{r}
l <- list(1, "a", TRUE, 1+1i)
names(l) <- c('entry1', 'entry2', 'entry3', 'entry4')
l$entry1
l$entry3
```
##Matrices
Matrices are vectors with a dimension attribute.
```{r}
mat <- matrix(nrow = 3, ncol = 4) # create a matix of dimension 3x4
mat
attributes(mat) # The dim attribute is itself a vector of lenght 2
dim(mat) # Get the dimension of the matrix
```
###Matrix Creation
Matrices are constructed by filling it column by column.
```{r}
mat <- matrix(1:12, nrow = 3, ncol = 4) 
mat
mat <- matrix(c('a', 'b', 'c', 'd'), nrow = 2, ncol = 2) 
mat
```
Or by using rbind/cbind.
```{r}
cbind(1:3, 6:8) # create a matrix by binding two column vectors together
rbind(1:3, 6:8) # create a matrix by binding two row vectors together
```
###Indexing
Matrices can be accessed with [row, column] 
```{r}
mat <- matrix(c('a', 'b', 'c', 'd'), nrow = 2, ncol = 2)
mat
mat[1,1]
mat[1,2]
```
##Data Frames
Data frames are lists of vectors. They are used to store tabular data.
```{r}
df <- data.frame(nr = 1:3, name = c("John", "Joe", "Jamie")) # construct a simple dataframe
df
nrow(df) # number of rows
ncol(df) # number of columns
```
###Indexing
Dataframes can be accessed like matrices (with [row, column]) or like lists (with the $ operator).
```{r}
df[1,2]
df$name
df$name[1]
```
##Factors
Factors are used to represent categorical data.
```{r}
facts <- factor(c("big","medium","small","big", "small"))
facts
```
Factors can also be ordered.
```{r}
facts <- factor(c("big","medium","small","big", "small"),
                levels = c("small", "medium", "big")) # construct factors with specific label ordering
facts
```
##Automatic Coercion
If datatypes are R is going to convert them to the least common datatype.
```{r}
str(c(1,"a")) # 1 is converted to chr
str(c(TRUE, 1)) # TRUE is converted to num
```
##Manual Data Conversations
You can convert objects to different types with the as.* functions.
```{r}
vec <- c(1,2,3)
str(vec)
as.numeric(vec)
as.character(vec)
```
##Missing Values
Missing values are denoted by NA or NaN
```{r}
is.na(NA) # function to test for NA of an object
is.nan(NaN) # function to test for NaN of an object
is.na(NaN) # NaN are also NA
is.nan(NA) # but NA are not NaN
is.na(c(1, NA, 2)) # produce a logical vector that indicates wether an element is NA or not
is.nan(0/0)
```
#Some functions that help us to explore R objects
```{r}
vec <- c(1,2,3)
str(vec) # structure of an object 
length(vec) # number of elements or components
class(vec)  # class or type of an object
summary(vec) # different information about the object
attributes(matrix(1:2)) # print attributes of the object
table(c("a","a","b")) # interpret input data as factors and do a frequency count
```
#Vector Operations
R has more or less the standart set of Operators.
What is special, is that the operators work on vector.
```{r}
vec <- c(1,2,3)
vec
vec + 1
vec * 2
vec / 3
vec^2
```
By default, R makes point-wise operations (Not like MATLAB).
```{r}
vec <- c(1,2,3)
vec * vec
```
If we want a vector-multiplication we need a special operator
```{r}
vec <- c(1,2,3)
vec %*% vec
```
The same is true for matrices.
```{r}
mat <- matrix(1:4, ncol=2, nrow=2)
mat
mat^2
mat * mat
mat %*% mat
```
##Recycling
When the vector arguments do not fit, R recycles the shorter vector.
```{r}
vec4 <- c(1,2,3,4)
vec2 <- c(1,2)
vec2 + vec4
vec4 + vec2
vec2 * vec4
```